A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-03 _tex/manuscript.tex
Built site for gh-pages

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we take the likelihood of observing
a patient in t-stage \(T_j\) and with involvement \(\xi_i\) to the power
of \(N_{ij}\), corresponding to the number of observations of the
patients in state \(i\) and t-stage \(T_j\) (either early or late) in
the dataset \(\mathbf{D}\).

@@

\[
\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
\]

@@

The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
are patient-specific and represent the assignment of each patient to a
particular model. We will later generalize patients with the same
diagnosis to the latent variable \(\boldsymbol{\epsilon}_{i,j}\). The
subscripts \(i\) and \(j\) represent that the latent variables are only
dependent on the diagnosis of the patient, which consists of lymphatic
involvement \(\boldsymbol{\xi}_i\) and t-stage \(T_j\). The latent
variables are central to the model, as they are hidden (unobserved) and
must be inferred.

@@

\begin{equation}\phantomsection\label{eq-complete-likelihood}{
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
}\end{equation}

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^s\) is the mixing parameter for subsite \(s\) and model \(m\),
\item
  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s data given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates whether
  patient \(k\) was generated by model \(m\).
\end{itemize}

@@

We can now apply the mixture model to our dataset \(\mathbf{D}\)
comprising the number of patients \(N_{ijs}\) diagnosed with LNL
involvement state \(\mathbf{X} = \boldsymbol{\xi}_i\), t-stage \(T_j\),
with a primary tumor in subsite \(s\).

@@

\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
}\end{equation}

In equation~\ref{eq-complete-likelihood} we compute the likelihood of
observing state \(\boldsymbol{\xi}_i\) with t-stage \(T_j\) under model
\(m\) similarly to equation~\ref{eq-likelihood-HMM}. However, we weight
each model with the mixing parameter \(\pi_m^s\). Since the mixing
parameters are subsite dependent, we differ between each subsite.

@@

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probabilities of a
patient \(k\) originating from one of the models \(m\). Given the
current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
the expectation value \(\gamma(\epsilon_m^k)\) is calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
}\end{equation}

THE OTHER OPTION

\begin{equation}\phantomsection\label{eq-expectation-step2}{
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
}\end{equation}

Here, the mixing parameter \(\pi_m^s\) depends on the subsite location
of the primary tumor of patient \(k\).

@@

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
}\end{equation}

@@

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
\]

@@

--\textgreater{} Add predictions and compare to the prevalence.
Additionally we could compare to tumor location predictions.

>>>>

\begin{equation}\phantomsection\label{eq-likelihood-HMM}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
}\end{equation}

In equation~\ref{eq-likelihood-HMM} we compute the likelihood of
observing patient the diagnosis of each patient \(k\), i.e.~t-stage
\(T_k\) and involvement \(\xi_k\).

@@

\[
\sum_{m}^M \pi_m^s = 1, \quad \forall s
\]

@@

The latent variables \(\boldsymbol{\epsilon}_{k}\), on the other hand,
are patient-specific and represent the assignment of each patient to a
particular model. The latent variables are central to the model, as they
are hidden (unobserved) and must be inferred.

@@

\begin{equation}\phantomsection\label{eq-complete-likelihood}{
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
}\end{equation}

@@

\begin{itemize}
\tightlist
\item
  \(\pi_m^{s_k}\) is the mixing parameter for patient \(k\) with a
  primary tumor in subsite \(s_k\) and model \(m\),
\item
  \(P(\mathbf{D}_k | \boldsymbol{\theta}_m)\) is the likelihood of
  patient \(k\)'s data given that it was generated by model \(m\),
\item
  \(\boldsymbol{\theta}_m\) represents the parameters of model \(m\),
\item
  \(\epsilon_{k}^m\) is the latent variable that indicates whether
  patient \(k\) was generated by model \(m\).
\end{itemize}

@@

@@

\begin{equation}\phantomsection\label{eq-complete-likelihood-explicite}{
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)} 
}\end{equation}

In equation~\ref{eq-complete-likelihood} we compute the likelihood of
observing the patient's state \(\boldsymbol{\xi}_k\) with t-stage
\(T_k\) under model \(m\) similarly to equation~\ref{eq-likelihood-HMM}.
However, we weight each model with the mixing parameter \(\pi_m^{s_k}\).
Since the mixing parameters are subsite dependent, we differ between
each subsite.

@@

In the E-step, we compute the expected value of the latent variables
\(\boldsymbol{\epsilon}\), which represent the probabilities of a
patient \(k\) originating from one of the models \(m\). Given the
current estimates of \(\boldsymbol{\theta}\) and \(\boldsymbol{\pi}\),
the expectation value \(\gamma(\epsilon_k^m)\) is calculated as:

\begin{equation}\phantomsection\label{eq-expectation-step}{
\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-maximization-step}{
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
}\end{equation}

@@

\[
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
\]

@@

add some analysis (level specific predictions and also comparison to
single tumor location models.)
